\large How resource abundance and stochasticity affect organisms' range sizes

### 2024-05-30 writing/manuscript.Rmd
updated writing

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
# author:
#   - name: Stefano Mezzini
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
#     correspondence: "yes"
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
# needs to be <= 200 words for AmNat
abstract <- "The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike. In this study, we leverage the available literature to provide a unifying framework and hypotheses for the effect of mean and variance in resources on organisms' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
```

@@

\noindent \textbf{Figures}: 6

@@

\noindent \textbf{References}: 95 <!-- counted by using nature csl -->

@@

\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`

@@

```{r wrap-figure-function, include = FALSE}
# this chunk is needed to wrap text around or beside figures
# code modified from https://stackoverflow.com/questions/54100806/r-markdown-how-do-i-make-text-float-around-figures

defOut <- knitr::knit_hooks$get('plot')  # save the default plot hook

knitr::knit_hooks$set(plot = function(x, options) { # set the new plot hook...

  x <- defOut(x, options)  # first apply the default hook

  # if wrapfigure options are given, create the new opening string for the
  # wrapfigure environment using custom figure alignment and text width
  if(!is.null(options$wrapfigure)) {

    # options must be numeric, they cannot be characters
    wf <- sprintf('\\begin{wrapfigure}{%s}{%g\\textwidth}',
                  options$wrapfigure[[1]], # figure alignment
                  options$wrapfigure[[2]]) # caption width relative to text width
    x  <- gsub(pattern = '\\begin{figure}', # and replace the default one with it
               replacement = wf,
               x = x,
               fixed = TRUE) # pattern is a string, so evaluate as is
    x  <- gsub(pattern = '{figure}', # also replace the environment ending
               replacement = '{wrapfigure}',
               x = x,
               fixed = TRUE)
  }
  return(x)
}
)

# chuck options used to wrap figures:
## fig.align = "right":          align figures right
## out.width = "0.5\\textwidth": output width = 0.5 * the width of the text
## wrapfigure = list("R", 0.5)   wrap caption to right, caption width = 0.5 text width

#' place figures before the paragraph they are referred to so `wrapfigure`
#' (generally) places them beside the paragraph
```

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018]. Thus, many organisms adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.<!-- Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, --> Movement represents one of the most readily available behaviors that species can adjust to optimize their resource use [<!--@krivan_ideal_2008; @brown_ecology_1999; -->@kacelnik_ideal_1992; @charnov_optimal_1976].<!-- whether this be by modifying their home range [@hirt_environmental_2021; @bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].-->

The relationship between organisms' movement and resource abundance has been of interest to biologists for nearly a century. @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how organisms decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., fig. 2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work. Instead, results are typically reported as independent, novel findings. Thus, there remains a need for a clear hypothesis for the effects of resource abundance on organisms' space-use requirements. Additionally, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources [but see: @stephens_optimal_1982; @rizzuto_forage_2021].

Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, herbivores in a grassland with relatively low but constant forage availability (low mean and variance) will adopt different behaviors and adaptations from herbivores in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (low mean and high stochasticity). In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency whinith patches [@charnov_optimal_1976]. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in different selection of movement and life history strategies. Although resource unpredictability (and seasonality) may ecrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource abundance on organisms' space use while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource stochasticity on organisms' space use while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics. We assume readers are familiar with fundamental statistical concepts (e.g., random variables, probability distributions, expected value and variance of a random variable), but we define and explain the necessary concepts in Appendix A.

<!--
\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix A.
-->

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., fig. 2 in the work of @bista_effect_2022]. Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime and a minimum amount of space it requires to survive [which is necessarily greater than zero -- see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].

```{r mu-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high space-use requirements as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, space-use requirements should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive. Note that the relationship between $\\e(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth'}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses: When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. At lower values of $\e(R)$, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (fig. A1).<!-- We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant.--> Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $\e(R)$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017] but will still not be able to predict $R$ exactly.

<!-- This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case. Most often, $\e(R)$ will vary over time, and organisms' behaviors will change in response. -->

@@

\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement. However, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant. Consequently, models estimating the effects of $R$ on organism's movement should account for not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant. Thus, we expect $\var(R)$ to have a stronger effect on space-use requirements when $\e(R)$ is low, and less of an effect when $\e(R)$ is high. We explore this more in the following section.

<!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy. However, the effects of $\var(R)$ on organisms' space-use requirements likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020]. Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small organisms, while organisms with larger space-use requirements may not even notice a change. -->

@@

\noindent The simulations in the section above mostly follow the hypotheses we presented in the introduction, but they are based on assumptions that are often not met in real natural enviroments. Organisms live in spatiotemporally heterogeneous and dynamic environments, which the development of perceptual ranges, navigation, and memory. Together, these abilities result in selective space use that depends on resource availability [@kacelnik_ideal_1992] and resource depletion [@charnov_optimal_1976].

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' range sizes
# author:
#   - name: Stefano Mezzini
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
#     correspondence: "yes"
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
# csl: 'the-american-naturalist-2024-03.csl'
# csl: 'nature.csl' # to count refs
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
# needs to be <= 200 words for AmNat
abstract <- "The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of an organism's range size, which challenges ecological theory and applied conservation alike. In this study, we leverage the available literature to provide a unifying framework and hypotheses for the effect of mean and variance in resources on range sizes. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in range size. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
```

@@

\noindent \textbf{Figures}: 5

@@

\noindent \textbf{References}: 109 (updated on 2024-05-29) <!-- counted by using nature csl -->

@@

\noindent \textbf{Key words:} energetics, environmental stochasticity, home range, range size, `ctmm`, simulations

@@

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms adapt their behaviors and/or physiology in response to changes in local resource abundance to ensure their needs are met.

While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976]. The relationship between organisms' movement and resource abundance has long been of interest to biologists. In his seminal paper, @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

```{r, echo=FALSE, eval=FALSE}
# visualizing the relationship
r <- seq(0.5, 5, by = 0.01)
her <- function(R, C = 1) return(C / R)
plot(r, her(R = r, C = 5), col = 'white', xlab = 'E(R)', ylab = 'E(H)',
     ylim = c(0, max(her(R = r, C = 4))))
for(i in 1:4) {
  lines(r, her(R = r, C = c(1, 2, 3, 4)[i]), col = i, lwd = 2)
}
```

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis. Many researchers have since demonstrated that organisms adapt their range sizes in response to resources abundance, but results are typically reported as independent, novel findings. Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021]. Thus, there remains a need for a clear hypothesis for the effects of both resource abundance and stochasticity on organisms’ range sizes.

Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. To see this, consider,, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance). This individual will adopt a different behavior and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity). Although these two habitats may have the same long-term average resource abundance, the ephemerality of resources in the stochastic desert environment would promote a different selection of movement and life history strategies. In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019]. Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [e.g., @chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

In this paper, we illustrate how an organism's range size can be expected to depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource abundance on range sizes while suggesting a simple and unifying hypothesis. Afterwards, we provide a review of the effects of resource stochasticity on organisms' range sizes while suggesting a second simple and unifying hypothesis. Subsequently, we support the two hypotheses using quantitative, simulated responses in range size to changes in resource abundance and stochasticity. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004]. Thus, we can treat the amount of resources $R$ at a given point in space ($u$) and time ($t$) as a random variable, denoted as $R(u, t)$. Treating resources as a random variable allows us to leverage techniques from probability theory and statistics, such as the expectation of a random variable (i.e., its mean) and its variance around the mean. We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, u)$ and $\sigma^2(t, u)$ to indicate them as functions of time ($t$) and space ($u$). Additional definitions and explanations are available in Appendix A.

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for the growth, survival, and reproduction of most organisms. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. As the abundance of essential resources fluctuates, motile organisms can move to new locations or 'patches' to meet their requirements [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for costs of movement [@taylor_energetics_1982].

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses, which is similar to that presented by @harestad_home_1979. When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. Like @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$. These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

```{r mu-hyp, fig.cap="Hypothesized range size of an organism as a function of resource abundance, $\\e(R)$. We expect low values of $\\e(R)$ to result in a large range, since organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident, nomadic, or migratory. As $\\e(R)$ increases, range size should decrease nonlinearly until it reaches the minimum amount of space required by the organism to survive. Note that the relationship between $\\e(R)$ and range size cannot be linear because it would require range size to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth'}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

Overall, the hypothesis that range size decreases with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because, conceptually, the relationship between range size and $\e(R)$ must be nonlinear, since: (1) there is an upper limit to how much space an organism is able to explore in its finite lifetime and (2) the minimum amount of space it requires to survive is necessarily greater than zero [see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above]. Consequently, we suggest analysts use models that account for this nonlinearity when estimating the effects of resource abundance on range size.

@@

\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ should be expected to affect organisms' range size, but we presented the two parameters as independent drivers of movement. However, organisms may respond to changes in $\sigma^2(t, u)$ more when resources are scarce than when they are abundant. Consequently, an organism’s movement behavior is likely to be a function of not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant. Thus, we expect $\var(R)$ to have a stronger effect on range size when $\e(R)$ is low, and less of an effect when $\e(R)$ is high. We explore this more in the following section.

@@

\noindent The simulations in the section above support the hypotheses we presented in the introduction, but they are based on assumptions that are often not met in real natural environments. Organisms live in spatiotemporally heterogeneous and dynamic environments that promote the use of perceptual ranges, navigation, and memory. Together, these abilities result in selective space use that depends on resource availability [@kacelnik_ideal_1992] and resource depletion [@charnov_optimal_1976].
